Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-06-29 MS/ms.Rmd
Just correcting spelling errors as I have a look at Shin's comments

In the words of Gene Glass, the father of meta-analysis: "Of course it mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest and the population one wishes to make conclusions about. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. <!---  as said, comparative analysis by definition, compares applies and oranges - = different species - I guess you are mainly talking about mixing different methods ones - this is where I said earlier - focusing on explaining hetero comes in. Also, mixing different quality of studies are often referred to as "Garbage in and garbage out" ---> Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis - an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a polluntant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. <!--- this is know as putting one number on so many different kinds of studies - but of course, this may not be that interesting apart from a starting point for subsequent meta-regressoin analyses ---> For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. It is important to consider these questions when conducting meta-analyses.

@@

As discussed above, it is clear that different types of questions and experimental approaches will be more or less susceptable to the 'apples and oranges' problem. <!--- not sure whehter it is a good idea to hightlight it as a problem as said earlier - explaining hetero is what's interesting and even the main porupose of meta-analysis for comparative physiologiests - they are not the problem but they are what we are going to invesitgate. I do not think 'suspetible' is a good word here given the average hetero = > 90% - you are nearly 100% guranteed to have hetero and this is what's interesting ---> As such, before embarking on our path to 'meta-analytic enlightenment', we conducted a literatrure survey to understand how widespread the problem may be in comparative physiology and gauge the ways in which it is currently being addressed, if at all. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptable to variation in how treatments are applied across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

@@

A very common occurance in comparative physiology is to manipulate the tempearture organisms experience and measure some physiologcal rate (e.g., metabolic rate). Using normal effect size measures (see above) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparablity' issues. As such, it is often of interest to standardsise effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

$$
\begin{equation} 
  lnRR_{Q_{10}} = ln\left( \frac{R_{2}}{R_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
(\#eq:lnq10)
\end{equation} 
$$ Equation \@ref(eq:lnq10) is essentially a temperature corrected equivalent to lnRR when the numerator and denomenator are measured at different temperatures. This allows one to compare the mean of two temperature treatments directly regardless of the temperatures that these groups have been measured. Here, we will refer to this as the log $Q_{10}$ response ratio, $lnRR_{Q_{10}}$. Notably, by taking the natural logarithm, there are two statistical advantages: 1) 0 becomes biological meaningful as 0 means two rates are the exactly the same and 'statistically significantly different from 0' means that two rates are significantly different to each other and 2) $lnRR_{Q_{10}}$ is more likely to satisfy the assumption of residual normality than $Q_{10}$ (see @Hedges1999). Also, the recognition of this equivalence means that we can easily calculate the sampling variance for equation \@ref(eq:lnq10) as follows:

@@

@Nakagawa2015 recently proposed analogous effect size estimates to *lnRR* that allow for comparisons of changes in variance between two groups, the log variance ratio (*lnVR*) and the log coefficient of variation (*lnCVR*). Like *lnRR*, *lnVR* and *lnCVR* are ratios that describe the relative difference in trait variablity between two groups. We refer readers to @Nakagawa2015 for the equations describing *lnVR* and *lnCVR*, but these can easily be extended to their $Q_{10}$ analogues (and associated sampling variances) as follows:

@@

It will not be possible to derive an effect size that completely controls for known heterogeneity at the effect size level all the time. This limitation is partly because treatment effects are not be simple linear or exponential functions of the mean response making the effect sizes we discuss above unsuitable [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships, with exponential or quadratic relationships in relation to some mean response. When this occurs we recommend comparative physiologists apply meta-regression approaches, controlling for these known treatment differences as a moderator. There are simple statisical approaches a meta-analyst can use to then make the overall effects interpretable [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytic approaches or steps commonly used.

@@

$$ Y_{i} = \beta_{0} + \sum \beta_{j}x_{j} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i} $$ where $\sum \beta_{j}x_{j}$ is simply the sum of all effects for all moderators (fixed effects), $x_{j}$. All other notation is the same as described above. Importantly, the variable $x_{j}$ could be any moderator collected from studies, including dosage, temperature, salinity, or pH. Some moderators will be at the level of the study (e.g., was the study experimental or observational in nature) but others could be taken at the effect size level (i.e., was the effect derived using a tempearture of 23°C or 35°C). By including these moderators in a meta-regression model, the interpretation of the overall meta-analytic mean, $\beta_{0}$, changes. For example, assume we collected the change in CORT dosage relative to a control group (i.e., in uL / g) for our meta-analysis on egg CORT effects on bird reproduction. If the untransformed dosage, $x_{j}$, is included in the model then the overall meta-analystic mean estimate, $\beta_{0}$, would be the mean effect of CORT relative to a control treatment, on reproduction when the dosage difference between the treatment and control group is 0 uL /g. Of course, this interpretation does not make much sense. We can, however, apply some simple transformations to make the overall mean estimate more intuitive to understand with the added benefit that it can help improve the interpretation of non-linear parameters estimated by the model [@Schielzeth2010; @GelmanHill2007]. One simple transformation would be to centre the variable $x_{j}$, by subtracting each value from the mean to create a new input variable (i.e., $c_{i}=x_{i}-\bar{x}$, where the subscript *i* denotes each individual effect size dosage). The new variable, $c_{j}$, is now centered on the mean and can replace $x_{j}$. When the model is refit with $c_{j}$ the estimated $\beta_{0}$ can now be more intuitivly interpreted as the overall mean effect of CORT on reproduction at an average dosage difference between treatment and control groups. Importantly, by centering, we retain the original units (uL / g) of the variable, which can be very useful to allow for a more general interpretation of the effect magnitude. Of course, if one is not interested in the magnitude of effect, we could also change our question and simply model mean reproductive output as a function of dosage (often what are referred to as 'arm-based' meta-analytic models). Here, one can model the mean for each group (i.e., control and treatment group), and acount for the known sampling variance associated with each group mean. Dosage, or other continuous variables like temperature, can then be modelled (accounting for sampling variance) to understand how mean reproductive output changes. It is important to recognise that all means need to be comparable and that these models are more complex to fit. While they can be equivalent to 'contrast-based' models, they require complex interactions to be estimated [@Nakagawa2015].

Centering moderator variables, like temperature and dosage, not only provides a way to make the overall mean effect more interpretable in the face of treatment heterogenity, but it also allows more complex relationships to be fit without compromising interpretations, including non-linear parameters and interactions (e.g., polynomials). In other words, linear / main effects can still be interpreted normally in the presence of non-linear / interactions fit in the model [@Schielzeth2010; @GelmanHill2007]. This can be particularly powerful when more complex, nonlinear, dosage relationships exist. We provide examples in the online supplementary material (with R code) to show readers how teh models we describe above can be fit, and how, meta-analytic means can be adjusted for treatment heterogenity.

@@

where $\sigma^2_{total} = \sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual} +\sigma^2_{m}$ is the total effect size variance (using the above multilevel model) and $\sigma^2_{m}$ is 'typical' the sampling error variance. <!--- you need more than that sigma_m is a typical one - you may need to provide a formula here --->. Quantifying heterogeneity provides a formal way to assess how much variation is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance) [@NakagawaSantos2012; @Higgins2002]. We can only conduct such analyses if we have an estimate of the sampling variance for a given effect measure because only then can we quantify how much variation is the result of sampling variability. @Senior2016 have shown that total heterogeneity is often extremely high in ecological and evolutionary meta-analysis (\~91%), suggesting that effect measures vary widely across and within studies. Some of the variation is clearly the result of working with different species or strains, different traits and different outcome measures, however, some of this variation also arises because of methodological differences across studies that might be related to the dosages or temperatures used for treatments. Formally quantifying and presenting heterogeneity estimates from meta-analytic models can provide a way to understand the major drivers of effect size variation by allowing researchers to produce various $I^2$ estimates which can be formally compared within and across studies. For example, we could fit the multilevel model described above to understand what proportion of variation is the result of, say, 'study' effects following @NakagawaSantos2012:

$$ I^2_{study} = \frac{\sigma^2_{study}}{\sigma^2_{total}} $$ where $\sigma^2_{study}$ is the study specific variance, and all other variance components are for phylogeny, species, residual (effect size level) and m (total measurement/sampling error). Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010]. If we want to formally assess how much effect size variation is the result of methodological factors, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects following @Nakgawa2013.

$$ R^2_{marginal} = \frac{\sigma^2_{fixed}}{\sigma^2_{fixed} + \sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual}} $$

>>>>

In the words of Gene Glass, the father of meta-analysis: "Of course it mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest and the population one wishes to make conclusions about. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. <!---  as said, comparative analysis by definition, compares applies and oranges - = different species - I guess you are mainly talking about mixing different methods ones - this is where I said earlier - focusing on explaining hetero comes in. Also, mixing different quality of studies are often referred to as "Garbage in and garbage out" ---> Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis - an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. <!--- this is know as putting one number on so many different kinds of studies - but of course, this may not be that interesting apart from a starting point for subsequent meta-regressoin analyses ---> For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. It is important to consider these questions when conducting meta-analyses.

@@

As discussed above, it is clear that different types of questions and experimental approaches will be more or less susceptable to the 'apples and oranges' problem. <!--- not sure whehter it is a good idea to hightlight it as a problem as said earlier - explaining hetero is what's interesting and even the main porupose of meta-analysis for comparative physiologiests - they are not the problem but they are what we are going to invesitgate. I do not think 'suspetible' is a good word here given the average hetero = > 90% - you are nearly 100% guranteed to have hetero and this is what's interesting ---> As such, before embarking on our path to 'meta-analytic enlightenment', we conducted a literature survey to understand how widespread the problem may be in comparative physiology and gauge the ways in which it is currently being addressed, if at all. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to variation in how treatments are applied across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

@@

A very common occurrence in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using normal effect size measures (see above) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparability' issues. As such, it is often of interest to standardsise effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

@@

$$
\begin{equation} 
  lnRR_{Q_{10}} = ln\left( \frac{R_{2}}{R_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
(\#eq:lnq10)
\end{equation} 
$$ Equation \@ref(eq:lnq10) is essentially a temperature corrected equivalent to lnRR when the numerator and denominator are measured at different temperatures. This allows one to compare the mean of two temperature treatments directly regardless of the temperatures that these groups have been measured. Here, we will refer to this as the log $Q_{10}$ response ratio, $lnRR_{Q_{10}}$. Notably, by taking the natural logarithm, there are two statistical advantages: 1) 0 becomes biological meaningful as 0 means two rates are the exactly the same and 'statistically significantly different from 0' means that two rates are significantly different to each other and 2) $lnRR_{Q_{10}}$ is more likely to satisfy the assumption of residual normality than $Q_{10}$ (see @Hedges1999). Also, the recognition of this equivalence means that we can easily calculate the sampling variance for equation \@ref(eq:lnq10) as follows:

@@

@Nakagawa2015 recently proposed analogous effect size estimates to *lnRR* that allow for comparisons of changes in variance between two groups, the log variance ratio (*lnVR*) and the log coefficient of variation (*lnCVR*). Like *lnRR*, *lnVR* and *lnCVR* are ratios that describe the relative difference in trait variability between two groups. We refer readers to @Nakagawa2015 for the equations describing *lnVR* and *lnCVR*, but these can easily be extended to their $Q_{10}$ analogues (and associated sampling variances) as follows:

@@

It will not be possible to derive an effect size that completely controls for known heterogeneity at the effect size level all the time. This limitation is partly because treatment effects are not be simple linear or exponential functions of the mean response making the effect sizes we discuss above unsuitable [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships, with exponential or quadratic relationships in relation to some mean response. When this occurs we recommend comparative physiologists apply meta-regression approaches, controlling for these known treatment differences as a moderator. There are simple statistical approaches a meta-analyst can use to then make the overall effects interpretable [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytic approaches or steps commonly used.

@@

$$ Y_{i} = \beta_{0} + \sum \beta_{j}x_{j} + a_{k[i]} + sp_{k[i]} + s_{j[i]} + e_{i} + m_{i} $$ where $\sum \beta_{j}x_{j}$ is simply the sum of all effects for all moderators (fixed effects), $x_{j}$. All other notation is the same as described above. Importantly, the variable $x_{j}$ could be any moderator collected from studies, including dosage, temperature, salinity, or pH. Some moderators will be at the level of the study (e.g., was the study experimental or observational in nature) but others could be taken at the effect size level (i.e., was the effect derived using a tempearture of 23°C or 35°C). By including these moderators in a meta-regression model, the interpretation of the overall meta-analytic mean, $\beta_{0}$, changes. For example, assume we collected the change in CORT dosage relative to a control group (i.e., in uL / g) for our meta-analysis on egg CORT effects on bird reproduction. If the untransformed dosage, $x_{j}$, is included in the model then the overall meta-analytic mean estimate, $\beta_{0}$, would be the mean effect of CORT relative to a control treatment, on reproduction when the dosage difference between the treatment and control group is 0 uL /g. Of course, this interpretation does not make much sense. We can, however, apply some simple transformations to make the overall mean estimate more intuitive to understand with the added benefit that it can help improve the interpretation of non-linear parameters estimated by the model [@Schielzeth2010; @GelmanHill2007]. One simple transformation would be to centre the variable $x_{j}$, by subtracting each value from the mean to create a new input variable (i.e., $c_{i}=x_{i}-\bar{x}$, where the subscript *i* denotes each individual effect size dosage). The new variable, $c_{j}$, is now centered on the mean and can replace $x_{j}$. When the model is refit with $c_{j}$ the estimated $\beta_{0}$ can now be more intuitively interpreted as the overall mean effect of CORT on reproduction at an average dosage difference between treatment and control groups. Importantly, by centering, we retain the original units (uL / g) of the variable, which can be very useful to allow for a more general interpretation of the effect magnitude. Of course, if one is not interested in the magnitude of effect, we could also change our question and simply model mean reproductive output as a function of dosage (often what are referred to as 'arm-based' meta-analytic models). Here, one can model the mean for each group (i.e., control and treatment group), and account for the known sampling variance associated with each group mean. Dosage, or other continuous variables like temperature, can then be modeled (accounting for sampling variance) to understand how mean reproductive output changes. It is important to recognise that all means need to be comparable and that these models are more complex to fit. While they can be equivalent to 'contrast-based' models, they require complex interactions to be estimated [@Nakagawa2015].

Centering moderator variables, like temperature and dosage, not only provides a way to make the overall mean effect more interpretable in the face of treatment heterogeneity, but it also allows more complex relationships to be fit without compromising interpretations, including non-linear parameters and interactions (e.g., polynomials). In other words, linear / main effects can still be interpreted normally in the presence of non-linear / interactions fit in the model [@Schielzeth2010; @GelmanHill2007]. This can be particularly powerful when more complex, nonlinear, dosage relationships exist. We provide examples in the online supplementary material (with R code) to show readers how the models we describe above can be fit, and how, meta-analytic means can be adjusted for treatment heterogeneity.

@@

where $\sigma^2_{total} = \sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual} +\sigma^2_{m}$ is the total effect size variance (using the above multilevel model) and $\sigma^2_{m}$ is 'typical' the sampling error variance. <!--- you need more than that sigma_m is a typical one - you may need to provide a formula here --->. Quantifying heterogeneity provides a formal way to assess how much variation is the result of 'true' biological or methodological differences as opposed to simply chance (i.e., sampling variance) [@NakagawaSantos2012; @Higgins2002]. We can only conduct such analyses if we have an estimate of the sampling variance for a given effect measure because only then can we quantify how much variation is the result of sampling variability. @Senior2016 have shown that total heterogeneity is often extremely high in ecological and evolutionary meta-analysis (\~91%), suggesting that effect measures vary widely across and within studies. Some of the variation is clearly the result of working with different species or strains, different traits and different outcome measures, however, some of this variation also arises because of methodological differences across studies that might be related to the dosages or temperatures used for treatments. Formally quantifying and presenting heterogeneity estimates from meta-analytic models can provide a way to understand the major drivers of effect size variation by allowing researchers to produce various $I^2$ estimates which can be formally compared within and across studies. For example, we could fit the multilevel model described above to understand what proportion of variation is the result of, say, 'study' effects following @NakagawaSantos2012: $I^2_{study} = \frac{\sigma^2_{study}}{\sigma^2_{total}}$ where $\sigma^2_{study}$ is the study specific variance, and all other variance components are for phylogeny, species, residual (effect size level) and m (total measurement/sampling error). Many readers will notice the similarities between $I^2$ and $R^2$ [@Nakgawa2013; @Nakagawa2010]. If we want to formally assess how much effect size variation is the result of methodological factors, such as temperature and dosage differences applied across studies, then we could fit our multilevel meta-regression model, including dosage or temperature (linear, quadratic or even more complex fits), and estimate how much variation is explained by these fixed effects following @Nakgawa2013.

$$ R^2_{marginal} = \frac{\sigma^2_{fixed}}{\sigma^2_{fixed} + \sigma^2_{study} + \sigma^2_{phylogeny} + \sigma^2_{species} + \sigma^2_{residual}}$$
